WASP-4 is Accelerating Toward the Earth

### 2020-02-28 paper/ms.tex
minor
update word

%
% contributing authors: alphabetical
%
\author[0000-0001-8638-0320]{A. W. Howard}
\affiliation{Cahill Center for Astrophysics, California Institute of
Technology, Pasadena, CA 91125, USA}
%
\author{S. B. Howell}
\affiliation{NASA Ames Research Center, Moffett Field, CA 94035, USA}
%
\author[0000-0002-0531-1073]{H. Isaacson}
\affiliation{Astronomy Department, University of California, Berkeley,
CA 94720, USA}
%
\author{H. Knutson}
\affiliation{Division of Geological and Planetary Sciences, California
Institute of Technology, Pasadena, CA 91125, USA}
%
\author[0000-0001-7233-7508]{R. A. Matson}
\affiliation{U.S. Naval Observatory, Washington, DC 20392, USA}
%

\begin{abstract}
  The orbital period of the hot Jupiter WASP-4b appears to be
  decreasing at a rate of $-8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$,
  based on transit-timing measurements spanning 12 years.  Proposed
  explanations for the period change include tidal orbital decay,
  apsidal precession, and acceleration of the system along the line of
  sight.  To investigate further, we performed new radial velocity
  measurements and speckle imaging of WASP-4.  The radial-velocity
  data show that the system is accelerating towards the Sun at a rate
  of $-0.0422\pm 0.0028\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.  The
  associated Doppler effect should cause the apparent period to shrink
  at a rate of $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$, comparable
  to the observed rate.  Thus, the observed change in the transit
  period is mostly or entirely produced by the line-of-sight
  acceleration of the system.  This acceleration is probably caused by
  a wide-orbiting companion of mass 10-200$\,M_{\rm Jup}$ and orbital
  distance 10-100$\,$AU, based on the magnitude of the radial-velocity
  trend and the non-detection of any companion in the speckle images.
  We expect that the orbital periods of 1 out of 6 hot Jupiters will
  appear to shrink at a rate similar to WASP-4b, based on the
  hot-Jupiter companion statistics of \citet{knutson_friends_2014}.
  Continued radial velocity monitoring of hot Jupiters is therefore
  essential to distinguish the effects of tidal orbital decay or
  apsidal precession from line-of-sight acceleration.
\end{abstract}

@@

Based on a priori expectations, if the outer companion lives within
100$\,$AU, then it seems more likely to be a brown-dwarf than a
low-mass star.  The reason is that stellar companions within 100$\,$AU
seem to be rare in systems with transiting planets, relative to field
stars.  This issue has been reviewed by \citet{moe_impact_2019}, who
synthesized work by \citet{wang_influence_2014,wang_influence_2015},
\citet{kraus_impact_2016}, \citet{ngo_2015_fohj2}, \citet{ngo_friends_2016},
\citet{matson_stellar_2018}, \citet{ziegler_soar_2020}, and others.
A handful of systems with inner hot Jupiters and outer brown dwarfs within
100$\,$AU are known, for instance CoRoT-20, HATS-59, WASP-53,
and WASP-81
\citep{triaud_peculiar_2017,rey_brown_2018,sarkis_hats-59bc_2018}.  In
contrast, from surveys by
\citet{knutson_friends_2014}, \citet{ngo_2015_fohj2} and
\citet{mugrauer_search_2019} we could find only one example of a
system with an inner hot Jupiter and an outer low-mass star within
100$\,$AU: HAT-P-10 \citep{bakos_2009_hatp10}.  Overall, this body of
literature indicates that the presence of a stellar-mass companion
within 100$\,$AU of the host star could hinder the formation of
planetary systems.  If true, then one would expect that the outer
companion in WASP-4 would have the lowest mass allowed by the data.

>>>>

%
% contributing authors: alphabetical
%
\author[0000-0001-8638-0320]{A. W. Howard}
\affiliation{Cahill Center for Astrophysics, California Institute of
Technology, Pasadena, CA 91125, USA}
%
\author[0000-0002-2532-2853]{S. B. Howell}
\affiliation{NASA Ames Research Center, Moffett Field, CA 94035, USA}
%
\author[0000-0002-0531-1073]{H. Isaacson}
\affiliation{Astronomy Department, University of California, Berkeley,
CA 94720, USA}
%
\author{H. Knutson}
\affiliation{Division of Geological and Planetary Sciences, California
Institute of Technology, Pasadena, CA 91125, USA}
%
\author[0000-0001-7233-7508]{R. A. Matson}
\affiliation{U.S. Naval Observatory, Washington, DC 20392, USA}
%

\begin{abstract}
  The orbital period of the hot Jupiter WASP-4b appears to be
  decreasing at a rate of $-8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$,
  based on transit-timing measurements spanning 12 years.  Proposed
  explanations for the period change include tidal orbital decay,
  apsidal precession, and acceleration of the system along the line of
  sight.  To investigate further, we performed new radial velocity
  measurements and speckle imaging of WASP-4.  The radial-velocity
  data show that the system is accelerating towards the Sun at a rate
  of $-0.0422\pm 0.0028\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.  The
  associated Doppler effect should cause the apparent period to shrink
  at a rate of $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$, comparable
  to the observed rate.  Thus, the observed change in the transit
  period is mostly or entirely produced by the line-of-sight
  acceleration of the system.  This acceleration is probably caused by
  a wide-orbiting companion of mass 10-200$\,M_{\rm Jup}$ and orbital
  distance 10-100$\,$AU, based on the magnitude of the radial-velocity
  trend and the non-detection of any companion in the speckle images.
  We expect that the orbital periods of 1 out of 3 hot Jupiters will
  change at rates similar to WASP-4b, based on the hot-Jupiter
  companion statistics of \citet{knutson_friends_2014}.
  Continued radial velocity monitoring of hot Jupiters is therefore
  essential to distinguish the effects of tidal orbital decay or
  apsidal precession from line-of-sight acceleration.
\end{abstract}

@@

Based on a priori expectations, if the outer companion lives within
100$\,$AU, then it seems more likely to be a brown-dwarf than a
low-mass star.  The reason is that stellar companions within 100$\,$AU
seem to be rare in systems with transiting planets, relative to field
stars.  This issue has been reviewed by \citet{moe_impact_2019}, who
synthesized work by \citet{wang_influence_2014,wang_influence_2015},
\citet{ngo_2015_fohj2}, \citet{ngo_friends_2016}, \citet{kraus_impact_2016},
\citet{matson_stellar_2018}, \citet{ziegler_soar_2020}, and others.
A handful of systems with inner hot Jupiters and outer brown dwarfs within
100$\,$AU are known, for instance CoRoT-20, HATS-59, WASP-53,
and WASP-81
\citep{triaud_peculiar_2017,rey_brown_2018,sarkis_hats-59bc_2018}.  In
contrast, from surveys by
\citet{knutson_friends_2014}, \citet{ngo_2015_fohj2} and
\citet{mugrauer_search_2019} we could find only one example of a
system with an inner hot Jupiter and an outer low-mass star within
100$\,$AU: HAT-P-10 \citep{bakos_2009_hatp10}.  Overall, this body of
literature indicates that the presence of a stellar-mass companion
within 100$\,$AU of the host star could hinder the formation of
planetary systems.  If true, then one would expect that the outer
companion in WASP-4 would have the lowest mass allowed by the data.
